### 2024-04-05 doc/15yrsChange.Rmd
discussion - minor changes from google doc

While nuisance species have had a large impact on AGB loss at our site (primarily through increased AWM of *Fraxinus americana*; Anderson-Teixeira et al, 2021), mortality among oak species, which is not linked to a non-indigenous nuisance species, is substantial as well.
Oak species in the region are generally found in older, even aged cohorts, which can contribute to accelerated mortality rates (Oak et al., 2016).
However, large oak mortality events, which often occur over protracted periods of time, have been common globally over the past century.
These mortality events are often termed “oak decline” events due to the complex abiotic and biotic factors which interact to incite mortality (Manion and Lachance, 1992).
Oak decline is believed to be the result of an imbalance in carbohydrate dynamics, which weakens oak trees’ resistance to pest and pathogen invasion, eventually leading to mortality (Manion and Lachance, 1992).

Relying on the current understory as a proxy for future canopy presents challenges.
White-tailed deer populations have varied greatly over the last century (citation).
Moreover, some canopy species, including some *Quercus* spp., are hypothesized to recruit in synchronized pulse events (Koenig, 2021).
A decline in deer population that reduces browsing pressure and coincides with a recruitment pulse, could allow for substantial, rapid regeneration of canopy species.
However, current recruitment levels are below a rate which we would expect in maturing secondary forests.
When overlaid with increasing mortality, regeneration may be insufficient to prevent extended periods without a closed canopy.

Despite inherent limitations in extrapolating findings from a single site, we find robust evidence suggesting broader applicability.
Mixed-hardwood forests, akin to our study site, are typical of the mid-Atlantic region. The persistently elevated deer populations documented at this site are reported throughout eastern US forests.
Furthermore,the widespread occurrence of vulnerable species (*e.g.* *Fraxinus* spp.) and continued spread of non-indigenous nuisance species (*e.g.* emerald ash borer, EAB) reinforces the relevance of our findings to the broader regional context.
EAB consistently induces near-total mortality in *Fraxinus* spp. once infected, to such an extent that five species in North America (including *Fraxinus americana*) are now critically endangered (Knight et al, 2013, IUCN Red list).

Our results underscore the importance of accounting for nuisance species in higher-level models and understanding of temperate forest ecosystem processes.
Nuisance species, driven by human introduction and perturbations to ecological equilibriums, interact with other anthropogenic changes to modify forest ecosystems.
By increasing mortality, nuisance species have the capacity to alter short-term carbon dynamics, resulting in large carbon losses that would not be expected in models that exclude them.
Through recruitment suppression, nuisance species can have long-term effects that may alter successional dynamics. They can delay recovery from disturbances, anthropogenic or natural, and even suppress regeneration into a closed-canopy for sustained periods.
They also can decrease resilience to other pressures as pests and pathogens decrease canopy diversity by shrinking the number of possible canopy species.

>>>>

While nuisance species have had a large impact on AGB loss at our site (primarily through increased AWM of *Fraxinus americana*; Anderson-Teixeira et al, 2021), mortality among oak species, which is not linked to a non-indigenous nuisance species, is substantial as well.
Oak species in the region are generally found in older, even aged cohorts, which can contribute to accelerated mortality rates (Oak et al., 2016).
However, large oak mortality events, which often occur over protracted periods of time, have been common globally over the past century.
These mortality events are often termed “oak decline” due to the complex abiotic and biotic factors which interact to incite mortality (Manion and Lachance, 1992).
Oak decline is believed to be the result of an imbalance in carbohydrate dynamics, which weakens oak trees’ resistance to pest and pathogen invasion, eventually leading to mortality (Manion and Lachance, 1992).

Relying on the current understory as a proxy for future canopy presents challenges.
White-tailed deer populations have varied greatly over the last century [@citation].
Moreover, some canopy species, including some *Quercus* spp., are hypothesized to recruit in synchronized pulse events (Koenig, 2021).
A decline in deer population that reduces browsing pressure and coincides with a recruitment pulse could allow for substantial, rapid regeneration of canopy species.
However, current recruitment levels are below a rate which we would expect in maturing secondary forests.
When overlaid with increasing mortality, regeneration may be insufficient to prevent extended periods without a closed canopy.

Despite inherent limitations in extrapolating findings from a single site, we find robust evidence suggesting broader applicability.
Mixed-hardwood forests, akin to our study site, are typical of the mid-Atlantic region. The persistently elevated deer populations documented at this site are reported throughout eastern US forests.
Furthermore, the widespread occurrence of vulnerable species (*e.g.* *Fraxinus* spp.) and continued spread of non-indigenous nuisance species (*e.g.* emerald ash borer, EAB) reinforces the relevance of our findings to the broader regional context.
EAB consistently induces near-total mortality in *Fraxinus* spp. once infected, to such an extent that five species in North America (including *Fraxinus americana*) are now critically endangered (Knight et al, 2013, IUCN Red list).

Our results underscore the importance of accounting for nuisance species in higher-level models and understanding of temperate forest ecosystem processes.
Nuisance species, driven by human introduction and perturbations to ecological equilibria, interact with other anthropogenic changes to modify forest ecosystems.
By increasing mortality, nuisance species have the capacity to alter short-term carbon dynamics, resulting in large carbon losses that would not be expected in models that exclude them.
Through recruitment suppression, nuisance species can have long-term effects that may alter successional dynamics. They can delay recovery from disturbances, anthropogenic or natural, and even suppress regeneration into a closed-canopy for sustained periods.
They also can decrease resilience to other pressures as pests and pathogens decrease canopy diversity by shrinking the number of possible canopy species.
